Species motif participation provides unique information about species risk of extinction

### 2022-05-24 manuscript/response_PRSB_round2.tex

\them
Second, as reviewer 2 argues there is scope for including a figure about the motifs in the context of their food web to facilitate the understanding about their structure and ecological meaning. The top of Figure 1 helps, but there is room to expand.

\us
Anna E

\them
Last, the authors generate what they argue are "a range of plausible network architectures".  But there is never any comparison of these architecture to empirical networks.  Nor is there any calculation of the key motif parameters on empirical data (for example the x-axis of figure 1). It will expand the audience for this manuscript greatly to include some empirical data to help empiricists understand for example in well-studied systems, what is the proportion of these different motifs.  This comparison does not need to be exhaustive, but it will serve two main roles in the manuscript: (1) provide support for the assertion that the authors have generated "a range of plausible network architectures" and (2) allow empiricists to understand how these simulations relate to real ecosystems.

This definitely sounds like the AE just wants a comparison of motif frequencies with empirical networks - much easier than showing same results!

Alyssa - this week (or bother me why not)
Webs referenced in the niche model vs empirical motifs paper

@@

\them
First, I think the introduction could be clearer. I think it spent too much time discussing bottom up effects, which I appreciate because that is one of the major limitations to this approach, however, I think some of the text could be better spent setting up some of the additional questions of the paper and clarifying various terminology (e.g. local, meso, and global scale gets thrown about, but is a bit of a surprise).

\us
The introduction has been re-written, and is hopefully clearer now regarding terminology and the questions asked. Additionally, we have now included a glossary in \emph{Appendix S11}

\them
Second, I found the analysis, particularly the divvying up into 5 parts rather circular. We know that connectance and species richness has a large impact on both motif expression and on persistence. Instead of multiple separate analyses, I think these can all be included in one large model.

\us [[May be related to some of the confusion about the purpose of different tests. Make sure the question asked by each analysis is clearer now.]] 
- Alyssa: S and C important but not main focus, so want to address them and then move on to more detailed things.
a written response here (and something brief in the paper)

@@

\them
Finally, I appreciate that the authors have used similar methods in a paper on BioRxiv – however, the current study relies quite heavily on it in the methodology. To my knowledge this paper is not peer-reviewed (or at least not yet) and thus I am less inclined to accept these citations as appropriate rationale for various decisions.

\us See above. The only methodological tie between the two manuscripts was when simulating the networks. The underlying software used is peer-reviewed~\citep{Delmas2017} and has been used in several other published studies, as well as the pre-print originally referenced (currently under review). To avoid the mis-perception that we rely on unproven methods, we now present the full network-generation methods in an Appendix.

\them
Smaller comments

Line 65 – 70: I agree that in-degree and trophic level may be coarse description of a species’ role within its food-web. However, the examples provided do not convince me that species’ roles based on motifs are a more nuanced way of describing a species’ role within its food web since the two examples provided refer to the influence disturbance may have on a basal resource – an influence that I’m not convinced a species’ role based on motifs will impact.

\us We have revised the introduction throughout, and have taken care to clearly link the extra information given by motifs to our intuition about extinction risk. In particular, motifs give information about a consumer's prey that may reflect the prey's risk of extinction and thereby the predators' risk (e.g., a predator consuming generalist prey will participate in many three-species chain motifs).

I do think that species’ roles based on motifs are a more nuanced way of describing species’ roles, and I am interested in the central question of this paper – however, this rationale would not convince me if I was unfamiliar with motifs/species roles/finer scale network descriptors.  As this is a central tenet of the paper I think the argument needs to be constructed better. Furthermore, just because a metric is coarse does not mean it is not an effective predictor – and when applying theoretical/computational results to empirical systems sometimes coarse is better. It is easier to measure the trophic position of a species using stable isotopes and approximate extinction probabilities than it is to construct an entire food web.

\us In our analysis, in-degree was a much stronger predictor of extinction risk than trophic level. We thus do not escape the need for full food-web information to predict extinction risk using coarse measures, even if we accept that stable isotope analyses are much less work than constructing a full food web (and whether stable isotopes are easier than constructing a full food web seems to largely depend on how many species are of interest ... we are not convinced that obtaining SI trophic levels for 100 species would be easier that estimating the full food web). Regardless, we are not suggesting that motifs replace degree or trophic level as all four metrics can be calculated using the same data. We are interested in showing that motifs can provide information about extinction risk that is not given by degree and trophic level.

[[Make sure conceptual figure shows how species with same degree, TL can participate in different motifs and have different risks - consumer with degree=2 consuming two specialists vs. consuming two generalists should make the point]]

@@

\them
Line 86: is this “focal” species the consumer mentioned in line 85?

\us
Rephrased, yes it is the same species, i.e. do the extinction risk of a consumer depend upon the motifs in which it participates in?

\them
Line 86: the impact of the motif profile of the full food web on the consumer’s extinction risk has come as a bit of a surprise to me. I think this could have been introduced better in the introduction – not extensively because I think it is probably a secondary question, but at this point I have no hypotheses for why I would even expect the motif profile of the full food web to out perform the motifs in which a focal species participates.

\us
We have now revised the introduction to introduce the motif profile of the whole network and its potential relevance to species-level extinction risk [[check - yes, I think so]] and hope that the purpose of these analyses is now clearer.
Note that we do not expect the frequencies of motifs at the network to out-perform species-level motif participation. In fact, we do not compare the predictive abilities of motif profiles and motif participation as these models relate to different things. We connect network-level motif profiles to the network-level average extinction risk of all consumers and species-level motif participation to species-level extinction risk of one consumer at a time. We regret the confusion about the goals of our different analyses in the previous draft and hope that it is now clear how our analyses fit together.

@@

\them
Line 87 – 89: similar to comment above – I think the idea of global and local food web properties has not been introduced at all, nor the classic food web properties (which ones are being measured? Only the four specific motifs thought to contribute to stability are mentioned).

\us We have revised the introduction to more clearly introduce these concepts. [[yes]]

\them
Line 92 – 95: this seems to be a flaw to the Bayesian network simulations – we know that mutual predation happens in food webs – especially if the food web isn’t explicitly size structured, and when nested within the broader food web context these feedback loops may have stabilizing, or destabilizing effects.

\us In revising the introduction, we now introduce the Bayesian network approach and its limitations in the methods section. [[might need to make this more explicit in the methods section, it is now mentioned a bit here and there, also add references to Allesina lab papers if we have some additional?]] [[I think it would be good to reiterate around this line that most secondary extinctions are captured well by the Bayesian framework. It does seem a little hard to believe that acyclic webs work so weel if you're new.]]

@@

\them
Line 182: what is network mean persistence? Species persistence is the probability of species persistence referred to in the methods? Be explicit. Also, what are the “other network properties”?

\us
Added clarifications to to point 3 and 5 where we had "network mean persistence" and "other network properties", please check if ok.

\them
Line 192: the idea of disturbance level came as a surprise – a careful reading of the methods pinpointed where you discuss this, however it is not discussed in the statistical analysis part, or in the main objectives of the paper at the end of the introduction.

\us This concept is now introduced in the introduction [[Disturbance is introduced, but different effects of different severities not clear yet.]] [[but should we discuss it a bit more also in the statistical analysis? we mention the different disturbance levels in the end of the methods section Calculating persistence]]

@@

\them
Line 193: Does it make sense to include Sn and Cn as random intercepts if it is known that both of these values impact 1) the number of motifs (which I don’t think has been rigorously controlled for – see above) and 2) network persistence? (same with eq’n S3)

\us We explicitly control for differences in the number of motifs through the frequency normalisation discussed above. We include the random intercepts to account for potential effects of Sn and Cn on network persistence that are \emph{not} due to effects on the total number of motifs. We agree with the Reviewer that the effects of Sn and Cn are well-known and important to account for, but they are not the main focus of our manuscript. We are therefore comfortable with keeping them as random effects in these equations.

@@

\them
Line 201: I have no idea what local and meso-scale refer to. Likewise global.

\us These terms are explained at their first use in the introduction or methods. To avoid confusion, we also add a glossary in \emph{Appendix S11} so that readers can easily look up these terms. We have also revised the paragraph starting on line 201 to include parenthetical definitions and clarify the scale which we are considering.

\them
Line 200-206: The statistical analyses of the various components of this study seem very circular to me. Various local metrics depend on global metrics but treated differently in each step. There must be a more appropriate statistical technique for considering all of these moving parts in a comprehensive manner – even using model selection to compare.

\us See response to XX ... probably need to repeat that our focus is not to throw away degree [[Definitely need to clarify our modelling framework]]

@@

\them
Line 23-26: this sentence could be clearer. Perhaps authors could rephrase to: “The structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole (Proulx et al., 2010; Scherber et al., 2010, Rosenblatt & Schmitz, 2016). For example, the loss of diversity…”

\us Changed as suggested.

\them
Line 57-59: This sentence seems counterintuitive to me. How could more connected networks be more resistant to secondary extinctions than less connected ones? A higher connectance indicates more pathways available where the energy flows, promoting the propagation of a disturbance (Vieira \& Almeida-Neto, 2015). Please justify your statement.
Vieira, M. C., \& Almeida‐Neto, M. (2015). A simple stochastic model for complex coextinctions in mutualistic networks: robustness decreases with connectance. Ecology Letters, 18(2), 144-152.

\us Alyssa: Expand explanation of why high connectance can be good sometimes (more prey available to switch to, albeit more disturbance pathways). Probably a good way to introduce size of disturbance - many pathways of a small disturbance may not matter, many pathways of a large more severe [[I think the current version does a succinct job of showing the good and bad of connectance]]

### 2022-05-24 manuscript/reorder_main_text.tex

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}. 
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. For example, a consumer with low trophic level may be at high risk of extinction if it consumes prey which are themselves very likely to be affected by disturbance,and a specialist may not necessarily have a high risk of extinction if it consumes a basal resource which is unlikely to be affected by disturbance.

    %\subsection*{3 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities, and provide information linking global-scale and local-scale properties, is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. 
    Such motif profiles of species are easily calculated once we know the network structure.
    Importantly, these `motif participation' roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a}. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}.

    This extra detail may be useful when predicting extinction risks of single species, but to our knowledge this has not yet been investigated. 
    However, at the network level, it has been shown that the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}.
    While these studies uncovered the potential influence of network-level motif frequencies on network stability, a disturbance may likely alter the network and the interactions between species in different ways. 
    This can affect the influence of motif frequencies on network stability, as it has been demonstrated that an unstable motif may appear stabilizing dependent on interactions with other species \citep{Kratina2012}.

>>>>

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, including secondary extinction due to reduced plant populations. 
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    As well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may, for example, be at greater risk of extinction than a specialist which consumes a highly generalist resource that is very unlikely to go extinct. 
    %For example, a consumer with low trophic level may be at high risk of extinction if it consumes prey which are themselves very likely to be affected by disturbance,and a specialist may not necessarily have a high risk of extinction if it consumes a basal resource which is unlikely to be affected by disturbance.

    %\subsection*{3 Motif participation may similarly affect species' particular extinction risk}
    It is possible to capture information about how a consumer and its prey fit together within a community by defining participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. 
    Such motif profiles of species are easily calculated once we know the network structure and require no more information than is used to calculate degree.
    These `motif participation' roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions~\citep{Cirtwill2015a,Cirtwill2018FoodWebs}. 
    By allowing researchers to distinguish between, for example, consumers of specialists and generalists, motif participation may complement degree and trophic level to more fully predict species' extinction risk.

    While the potential relationships between motif participation and species-level extinction risk have not yet been tested to our knowledge, it has been shown that the frequencies of different motifs within a whole network are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}.
    While these studies uncovered the potential influence of network-level motif frequencies on network stability, a disturbance may likely alter the network and the interactions between species in different ways. 
    This can affect the influence of motif frequencies on network stability, as it has been demonstrated that an unstable motif may appear stabilizing dependent on interactions with other species \citep{Kratina2012}.
